---
title: "Introduction to multibias"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to multibias}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(multibias)
```

Multibias makes it easy to simultaneously adjust for multiple biases in causal inference research. This document walks you through the key steps to performing the analysis.

## 1. Create data_observed

Represent your observed data as a `data_observed` object. Here you include the dataframe, specify the key variables in the data, and identify the bias impacting the data. All epidemiological biases can be generalized into four main groups:

* Uncontrolled confounding (**uc**)
* Exposure misclassificaiton (**em**)
* Outcome misclassificaiton (**om**)
* Selection bias (**sel**)

Multibias is capable of handling bias adjustment for most of the combinations of the above four biases.

For purposes of demonstration, multibias includes datasets with different bias combinations. For a given bias or biases, it includes the biased data (e.g., `df_uc_sel`: missing a confounder and not including the un-selected subjects) and the source used to derive the biased data (e.g., `df_uc_sel_source`: including data on the missing confounder and subjects who were not selected).

Each dataset has variables defined similarly:

* `X` = binary exposure
* `Y` = binary outcome
* `C1-C3` = binary confounders

```{r}
df_observed <- data_observed(
  df_uc_sel,
  bias = c("uc", "sel"),
  exposure = "X",
  outcome = "Y",
  confounders = c("C1", "C2", "C3")
)
```

## 2. Create source for bias adjustment

Next, you need to have some information that can be used to quantify the assumed bias or biases. There are two options here.

### Option 1: Bias parameters

One option is to list the assumed parameters for the corresponding bias equations. The necessary bias equations are provided in the documentation of each `adjust()` function. These equations can be derived using the methods outlined in the article [here](https://doi.org/10.1093/ije/dyad001). In our example, we have bias parameters to predict the missing binary confounder `U` (as the vector `u_coefs`) and study selection indicator `S` (as the vector `s_coefs`).

```{r}
bp <- bias_params(
  coef_list = list(
    u = c(-0.19, 0.61, 0.72, -0.09, 0.10, -0.15),
    s = c(-0.01, 0.92, 0.94)
  )
)
```

### Option 2: Validation data

A second option is to specify a validation data source as a `data_validation` object. In order to adjust for a given bias, the validation data must have the corresonding missing data. In the example here, the validation data has additional columns for the missing binary confounder `U` and indicator `S` for whether a given individual was selected into the study.

```{r}
df_validation <- data_validation(
  df_uc_sel_source,
  true_exposure = "X",
  true_outcome = "Y",
  confounders = c("C1", "C2", "C3", "U"),
  selection = "S"
)
```

## 3. Run the bias adjustment

Finally, we can run the multi-bias adjustment!

### Single run

```{r}
multibias_adjust(
  data_observed = df_observed,
  bias_params = bp
)
```

or

```{r}
multibias_adjust(
  data_observed = df_observed,
  data_validation = df_validation
)
```

We observe that the bias-adjusted odds ratio of the effect of X on Y is approximately 2. This effect can be compared to the effect observed in the biased data to understand whether systematic error is bringing the observed effect closer to the null or further from the null.

### Bootstrapping

To obtain a confidence interval that captures uncertainty in both the random error and the systematic error, bootstrapping is advised. The `multibias_adjust()` function comes with built-in parameters that make this simple. If improved computational performance is desired, users may want to incorporate parallelization.

```{r}
multibias_adjust(
  data_observed = df_observed,
  data_validation = df_validation,
  bootstrap = TRUE,
  bootstrap_reps = 10
)
```

When adjusting via bias parameters, the user can alternatively provide each bias parameter value as a probability distribution. The resulting confidence interval from bootstrapping can then quantify uncertainty in both the random error and the systematic error. Check out the vignette article "Multibias Validation" for a demonstration of this approach.
