---
title: "Multibias example: NHANES"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multibias example: NHANES}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Epidemiologic studies often face multiple sources of bias simultaneously, including confounding, measurement error, and selection bias. Traditional approaches typically address these biases one at a time, which can lead to incomplete adjustment and potentially misleading results. The `multibias` package provides a framework for simultaneously adjusting for multiple sources of bias.

This vignette demonstrates how to use `multibias` to analyze the relationship between alcohol consumption and mortality using data from the National Health and Nutrition Examination Survey (NHANES). We'll progressively add different types of bias adjustments to show how each affects our estimates and how they can be combined for a more comprehensive analysis.

## Setup

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(multibias)
library(dplyr)
```

For this demonstration, we'll use the 2013-2018 [National Health and Nutrition Examination Survey](https://www.cdc.gov/nchs/nhanes/about/index.html) (NHANES) data linked to the [CDC's 2019 public-use mortality data](https://www.cdc.gov/nchs/data-linkage/mortality-public.htm). This dataset provides a rich source of information including:

- Demographics
- 24-hour dietary recall
- Alcohol consumption patterns
- Smoking history
- Mortality outcomes

```{r}
nhanes <- read.csv("nhanes.csv")
```

## Data Preparation and Initial Exploration

We'll be assessing the risk of alcohol intake on all-cause mortality, specifically comparing "extreme" drinkers to "non-extreme" drinkers. Let's first understand our data and create our exposure variable.

### Creating the Exposure Variable

We'll define "extreme" drinking based on the [USDGA guidelines](https://www.dietaryguidelines.gov/alcohol/info) of:

- 14 g/day for women
- 28 g/day for men

Participants are classified as "extreme drinkers" if they reported drinking more than 1.5x these guidelines.

```{r}
nhanes_filtered <- nhanes |>
  mutate(alcohol_extreme = case_when(
    alcohol_day_total > 14 * 1.5 & gender_female == 1 ~ 1,
    alcohol_day_total > 28 * 1.5 & gender_female == 0 ~ 1,
    TRUE ~ 0
  )) |>
  filter(age >= 18) |>
  filter(eligstat == 1) |>
  filter(alcohol_12mo > 0)
```

### Data Characteristics

Let's examine the key characteristics of our dataset:

```{r}
exposure_outcome_table <- nhanes_filtered |>
  group_by(alcohol_extreme, mortstat) |>
  summarize(count = n()) |>
  ungroup() |>
  mutate(proportion = count / sum(count))

print("Distribution of Exposure and Outcome:")
print(exposure_outcome_table)

demographic_table <- nhanes_filtered |>
  group_by(alcohol_extreme) |>
  summarize(
    n = n(),
    age_mean = mean(age),
    age_sd = sd(age),
    gender_prop = mean(gender_female)
  ) |>
  mutate(
    alcohol_extreme = if_else(
      alcohol_extreme == 1, "Extreme", "Non-extreme"
    )
  )

print("\nDemographic Characteristics by Exposure Status:")
print(demographic_table)
```

## Base Model Analysis

Let's start with a basic logistic regression model to estimate the association between extreme alcohol consumption and mortality, adjusting for age and gender. This will serve as our reference point for comparing bias-adjusted estimates.

```{r}
base_mod <- glm(
  mortstat ~ alcohol_extreme + gender_female + age,
  data = nhanes_filtered,
  family = binomial(link = "logit")
)

base_results <- data.frame(
  term = c("Extreme Alcohol Use", "Female Gender", "Age"),
  estimate = round(exp(coef(base_mod)), 2)[-1],
  ci_lower = round(exp(confint(base_mod)), 2)[-1, 1],
  ci_upper = round(exp(confint(base_mod)), 2)[-1, 2]
)

print("Base Model Results: Odds Ratios and 95% Confidence Intervals")
print(base_results)
```

Our base model suggests that extreme alcohol consumption is associated with a 1.6-fold increase in mortality risk (95% CI: 1.1, 2.2), after adjusting for age and gender. However, this estimate may be biased due to several factors that we'll address in the following sections.

## Multiple Bias Analysis

We'll now progressively adjust for three types of bias:
1. Uncontrolled confounding (smoking status)
2. Exposure misclassification (alcohol underreporting)
3. Selection bias (NHANES sampling weights)

### 1. Adjusting for Uncontrolled Confounding

Smoking is a strong confounder in the relationship between alcohol consumption and mortality. Smokers are more likely to be heavy drinkers, and smoking independently increases mortality risk. Our base model didn't account for this, potentially leading to biased estimates.

To adjust for the bias here, we'll leverage a validation dataset; for a fraction of our cohort of patients, we have smoking information available.

```{r}
# Create observed data object with uncontrolled confounding bias
df_obs1 <- data_observed(
  data = nhanes_filtered,
  bias = "uc",
  exposure = "alcohol_extreme",
  outcome = "mortstat",
  confounders = c("age", "gender_female")
)

# Create validation data with smoking information
df_temp1 <- nhanes_filtered |>
  filter(!is.na(smoked_100cigs))

df_val1 <- data_validation(
  data = df_temp1,
  true_exposure = "alcohol_extreme",
  true_outcome = "mortstat",
  confounders = c("age", "gender_female", "smoked_100cigs")
)

# Perform bias adjustment
set.seed(1234)
uc_adjusted <- multibias_adjust(df_obs1, df_val1)
print("Uncontrolled Confounding Adjusted Results:")
print(uc_adjusted)
```

### 2. Adding Exposure Misclassification Adjustment

Self-reported alcohol consumption is often underreported due to social stigma. We'll adjust for this by assuming that certain demographic groups (wealthy, college-educated individuals) are more likely to under-report their alcohol consumption.

```{r}
# Create observed data object with both biases
df_obs2 <- data_observed(
  data = nhanes_filtered,
  bias = c("em", "uc"),
  exposure = "alcohol_extreme",
  outcome = "mortstat",
  confounders = c("age", "gender_female")
)

# Create validation data with adjusted alcohol consumption
df_temp2 <- nhanes_filtered |>
  filter(!is.na(smoked_100cigs)) |>
  mutate(
    # Assume 50% higher reported consumption for high SES individuals
    alcohol_adj = if_else(
      income == "$100,000 and Over" | education == "College graduate or above",
      alcohol_day_total * 1.5,
      alcohol_day_total
    )
  ) |>
  mutate(alcohol_extreme_adj = case_when(
    alcohol_adj > 14 * 1.5 & gender_female == 1 ~ 1,
    alcohol_adj > 28 * 1.5 & gender_female == 0 ~ 1,
    TRUE ~ 0
  ))

df_val2 <- data_validation(
  data = df_temp2,
  true_exposure = "alcohol_extreme_adj",
  true_outcome = "mortstat",
  confounders = c("age", "gender_female", "smoked_100cigs"),
  misclassified_exposure = "alcohol_extreme"
)

# Perform bias adjustment
set.seed(1234)
em_uc_adjusted <- multibias_adjust(df_obs2, df_val2)
print("Exposure Misclassification + Confounding Adjusted Results:")
print(em_uc_adjusted)
```

### 3. Adding Selection Bias Adjustment

NHANES uses a complex sampling design with oversampling of certain demographic groups. This can create selection bias if both the exposure (alcohol consumption) and outcome (mortality) are related to the selection probability. We'll adjust for this using the NHANES sampling weights.

```{r}
# Create observed data object with all three biases
df_obs3 <- data_observed(
  data = nhanes_filtered,
  bias = c("em", "uc", "sel"),
  exposure = "alcohol_extreme",
  outcome = "mortstat",
  confounders = c("age", "gender_female")
)

# Prepare data for selection bias adjustment
df_temp3a <- nhanes_filtered |>
  filter(!is.na(smoked_100cigs)) |>
  mutate(
    alcohol_adj = if_else(
      income == "$100,000 and Over" | education == "College graduate or above",
      alcohol_day_total * 1.5,
      alcohol_day_total
    ),
    alcohol_extreme_adj = case_when(
      alcohol_adj > 14 * 1.5 & gender_female == 1 ~ 1,
      alcohol_adj > 28 * 1.5 & gender_female == 0 ~ 1,
      TRUE ~ 0
    ),
    # Combine survey weights
    weight = if_else(weight_day2 == 0, weight_day1, weight_day2)
  )

# Create selection indicator based on sampling weights
set.seed(1234)
selected_sample <- sample(
  x = df_temp3a$seqn,
  size = nrow(df_temp3a),
  replace = TRUE,
  prob = df_temp3a$weight
)

df_selected_sample <- data.frame()
for (id in selected_sample) {
  df_selected_sample <- rbind(
    df_selected_sample,
    df_temp3a[df_temp3a$seqn == id, ]
  )
}

df_selected_sample$selection <- 1
df_not_selected_sample <- df_temp3a[!(df_temp3a$seqn %in% selected_sample), ]
df_not_selected_sample$selection <- 0

df_temp3b <- rbind(df_selected_sample, df_not_selected_sample)

# Create validation data with selection information
df_val3 <- data_validation(
  data = df_temp3b,
  true_exposure = "alcohol_extreme_adj",
  true_outcome = "mortstat",
  confounders = c("age", "gender_female", "smoked_100cigs"),
  misclassified_exposure = "alcohol_extreme",
  selection = "selection"
)

# Perform final bias adjustment
set.seed(1234)
final_adjusted <- multibias_adjust(df_obs3, df_val3)
print("All Biases Adjusted Results:")
print(final_adjusted)
```

## Comparing Results

Let's compare all our estimates to see how each bias adjustment affects our results:

```{r}
# Create comparison table
comparison_table <- data.frame(
  Model = c(
    "Base Model",
    "Adjusted for Uncontrolled Confounding",
    "Adjusted for Exposure Misclassification & Uncontrolled Confounding",
    "Adjusted for Three Biases"
  ),
  OR = c(
    round(exp(coef(base_mod))["alcohol_extreme"], 2),
    round(uc_adjusted$estimate, 2),
    round(em_uc_adjusted$estimate, 2),
    round(final_adjusted$estimate, 2)
  ),
  CI_lower = c(
    round(exp(confint(base_mod))["alcohol_extreme", 1], 2),
    round(uc_adjusted$ci[1], 2),
    round(em_uc_adjusted$ci[1], 2),
    round(final_adjusted$ci[1], 2)
  ),
  CI_upper = c(
    round(exp(confint(base_mod))["alcohol_extreme", 2], 2),
    round(uc_adjusted$ci[2], 2),
    round(em_uc_adjusted$ci[2], 2),
    round(final_adjusted$ci[2], 2)
  )
)

print("Comparison of Bias-Adjusted Estimates:")
print(comparison_table)
```

## Discussion

Our analysis demonstrates how multiple sources of bias can affect the estimated relationship between alcohol consumption and mortality:

1. **Uncontrolled Confounding**: Adjusting for smoking status reduced the odds ratio from 1.60 to 1.45, suggesting that part of the observed association was due to the confounding effect of smoking.

2. **Exposure Misclassification**: Further adjustment for potential underreporting of alcohol consumption reduced the odds ratio to 1.35, indicating that measurement error may have led to an overestimate of the true effect.

3. **Selection Bias**: The final adjustment for NHANES sampling weights had minimal impact on the estimate, suggesting that selection bias may not be a major concern in this analysis.

These results highlight the importance of considering multiple sources of bias simultaneously. The `multibias` package provides a flexible framework for such analyses, allowing researchers to incorporate various assumptions about bias mechanisms and assess their impact on study results.

## Limitations and Considerations

Several limitations should be considered when interpreting these results:

1. **Bias Assumptions**: Our adjustments rely on specific assumptions about bias mechanisms (e.g., 50% higher reported consumption among high SES individuals). These assumptions should be based on prior research and may need to be varied in sensitivity analyses.

2. **Validation Data**: The quality of bias adjustments depends on the availability and quality of validation data. In practice, such data may be limited or imperfect.

3. **Unmeasured Confounding**: While we adjusted for age, sex, and, smoking, other important confounders may exist that we couldn't account for.

4. **Generalizability**: NHANES participants may not be representative of the general population, even after accounting for sampling weights.

The `multibias` package allows researchers to easily test different assumptions and assess the sensitivity of their results to various bias scenarios.